Causal Discovery on Precarity and Depression

### 2025-02-11 draft_v1/draft_v1.qmd
update_results
only FCI in main text + proportion matrix plots

Mental health problems in urban areas have been reported to be on the rise the complexity of mental health systems presents significant challenges in understanding the underlying mechanisms driving these issues, let alone planning effective interventions.

Research has aimed to identify underlying factors contributing to mental health problems. Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper). This comprehensive perspective on precariousness helped to highlight how different aspects of life may be interconnected. While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health? Specifically, the lack of directional information — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue. Using causal discovery methods, we explore how different aspects of precariousness influence depression and delve deeper into the dynamics at the symptom level. By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to precariousness. Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.

@@

There are numerous causal discovery algorithms available; however, in this study, we focus on algorithms suited to the potential cyclic relationships within our system. Specifically, we use FCI (Fast Causal Inference) and CCI (Cyclic Causal Inference), both capable of accounting for such cycles under certain conditions [@mooij2020constraint; @strobl2019]. Additionally, we include the PC algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].

@@

As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions. Despite these differences, both graphs are forms of *ancestral graphs*, designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
Directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. The Markov equivalence class refers to the set of graphs that encode the same conditional independence relationships, such that the same *d-separation* relations hold across all graphs in the class [@spirtes2001causation]. 
$A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
When $A \arrowarrow B$, neither $A$ nor $B$ is an ancestor of the other, indicating the presence of a latent confounder between them.

The algorithms also differ in their approaches and assumptions for detecting cycles.
In CCI's MAAG, $A$ \textemdash\ $B$ indicates that $A$ is an ancestor of $B$ and $B$ is an ancestor of $A$, referring to a cycle between $A \stackedarrows B$. 
FCI, on the other hand, identifies potential cycles more subtly: fully-connected nodes with circle endpoints ($\circirc$) in FCI's PAG may suggest a cyclic structure. 
FCI basically provides a sufficient condition to distinguish nodes that are not part of a cycle, offering a more nuanced approach [@mooij2020constraint].
Table 1 highlights further differences in the assumptions underlying these algorithms. CCI operates under the assumption of a linear system, while FCI, particularly for inferring cyclic relationships, assumes a non-linear system without selection bias and adheres to more general faithfulness and Markov conditions (i.e., *$\sigma$-separation* and *$\sigma$-faithfulness* setting) [@forre2018constraint].
When the respective assumptions of each algorithm are satisfied, their outcomes for cyclic relationships align with those illustrated in @fig-examplegraphs.
While an exhaustive discussion of causal discovery concepts and algorithm-specific details is beyond the scope of this paper, readers seeking a deeper understanding are encouraged to consult @park2024discovering for comprehensive insights into these methods and their applications.

@@

The graph produced by the PC algorithm is a *CPDAG* (completed partially directed acyclic graph), where directed edges ($A$ → $B$) indicate that $A$ is a direct cause (parent) of $B$. Unlike FCI and CCI, the CPDAG does not include circle symbols. Instead, when the PC algorithm cannot determine the direction of causality, it represents this uncertainty with bidirectional arrows. While the PC algorithm serves as a useful reference, its strict assumptions of acyclicity and the absence of latent confounders restrict its applicability in more complex scenarios. Therefore, our primary focus remains on the results from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.

One practical challenge in applying these algorithms to the HELIUS dataset is that the data does not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear. To account for this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
While kernel-based conditional independence tests (KCIT) are effective, they are computationally intensive for large datasets like HELIUS, as their quadratic scaling with sample size is primarily driven by the inversion of large kernel matrices. To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximate kernel methods using random Fourier features. This approach reduces computation to linear scaling with sample size, significantly lowering computational costs [@strobl2019approximate]. For a more detailed explanation of RCoT, refer to @sec-rcot.

## Analysis
We analyze the causal structure using three approaches: examining the relationship (1) between the five individual precarity factors and the PHQ sum score, which represents overall depression severity; (2) between individual symptoms and precarity factors; and (3) between individual symptoms and overall precarity (the sum of the five precarity factors). 
Using the PHQ sum score simplifies the analysis by reducing dimensionality, offering computational efficiency and a broad, interpretable perspective on the relationship between depression and precarity factors. In contrast, analyzing individual symptoms provides a more detailed understanding by capturing the diverse ways symptoms respond to different precarity factors. However, this detailed approach introduces methodological challenges, including non-standard distributions of individual symptom variables and the complexities associated with increased dimensionality.
Finally, examining individual symptoms in relation to the overall precarity score complements the analysis by revealing how symptoms collectively connect to general precarity. Aggregating precarity factors may uncover relationships that could be overlooked in analyses of individual factors, particularly when a symptom has associations across multiple factors that are not strong enough to emerge in individual analyses.
By integrating these three approaches, we achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression. 
<!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions. -->

Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping. For each of the 100 bootstrap samples, we estimate causal graphs and retain only the edges and directions that exceed predefined thresholds.
The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms. Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm. For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
To further enhance efficiency, we fix the skeleton of edges among symptom variables based on the common structure estimated across all algorithms. This approach minimizes spurious edges in the symptom network and significantly reduces the computational time needed for skeleton estimation. 
This fixed structure also aligns well with commonly reported skeleton structures in the literature, ensuring consistency with existing findings (**cite the comp_model paper**).[^2]
Lastly, to summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.  See @fig-workflow for an overview of the analysis workflow.

[^2]:
We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.001. This approach preserved the complexity of symptom interactions, enabling unrestricted estimation of causal directions both among symptoms (*symptom --- symptom* interactions) and between symptoms and precarity factors.

[^3]:
The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix].

@@

# Results
## Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone. Black edges represent consistent causal relationships identified by both FCI and CCI, while gray dashed edges denote inconsistent relationships that vary between the two methods. The inconsistent endpoints of the gray edges are marked with circles.

In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*). *P.soc* does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder. Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*). Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.

Graph (b), derived solely from the nonparametric RCoT method. Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*. 
The relationship between stressors and depression remains unresolved between the two algorithms. However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.

@@

## Individual depression symptom
Moving from the sum score representation to the symptom-level graph provides a more granular perspective on the causal relationships between precarity factors and depression. This approach highlights the heterogeneity in how precarity factors influence individual depressive symptoms — *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood). While the sum score graph aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom graph uncovers distinct pathways for different symptoms.
In the symptom-level graph (@fig-sym), consistent relationships are represented by black solid edges, while areas of disagreement between the FCI and CCI algorithms are denoted by gray dashed edges. Endpoints marked with circles indicate differences in directional conclusions between the algorithms. Additionally, the navy dashed edges represent relationships unique to the graphs generated using Gaussian CI testing.

<!-- Overall Patterns in the Symptom Network -->
The symptom-level graph reveals a complex and interconnected structure, far more intricate than the sum score graph. Its denser network highlights the strong interdependence among symptoms and suggests the presence of latent confounding influences, as indicated by numerous bidirectional arrows.

<!-- Centrality and Connectivity -->
Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points. Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network. Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors. These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
<!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. -->

The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs. As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms. The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships. For instance, *slp* is identified as influencing *P.emp* or potentially through a latent confounder (*slp* o-> *P.emp*) , while *glt* appears to affect *P.soc* or may be linked via a confounder (*glt* o-> *P.soc*).
Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*. As seen in the sum graph, Gaussian CI testing tends to produce a denser graph. In this case, causal relationships involving *slp* are particularly prominent.

Some discrepancies, however, exist between the symptom-level and sum graphs. For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors. Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.

>>>>

Mental health problems in urban areas are on the rise, yet the complexity of mental health systems poses significant challenges in understanding the underlying mechanisms driving these issues — let alone designing effective interventions.
Research has long sought to identify underlying factors contributing to mental health problems. Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper). This comprehensive perspective on precariousness helped to highlight how different aspects of life may be interconnected. While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health? Specifically, the lack of directional insight — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue. 
Using causal discovery methods, we explore how various aspects of precariousness contribute to depression and whether depression, in turn, exacerbates precariousness, potentially creating a feedback loop.
Furthermore, we zoom in at the symptom level, examining individual depressive symptoms to identify those that may act as initiators — triggering the activation of the symptom network — by displaying heightened sensitivity to specific precariousness factors.
We also investigate symptoms that may function as bridging elements, linking the dynamics of depression with precariousness by feeding back into precarious conditions, potentially reinforcing the cycle of mental health deterioration and precarious conditions.
Through this analysis, we aim to uncover the causal mechanisms driving mental health challenges related to precariousness and establish a foundation for more targeted and effective interventions by identifying the key driving forces within this system.

@@

There are numerous causal discovery algorithms available; however, in this study, we focus on algorithms suited to the potential cyclic relationships within our system. Specifically, we use *FCI* (Fast Causal Inference) and *CCI* (Cyclic Causal Inference), both capable of accounting for such cycles [@mooij2020constraint; @strobl2019]. Additionally, we include the *PC* algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].

@@

As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions. 
Despite these differences, both graphs belong to the class of *ancestral graphs*, which are designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
In these graphs, directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. 
The Markov equivalence class represents a set of graphs that encode the same conditional independence relationships, ensuring that the same *d-separation* conditions hold across all graphs in the class [@spirtes2001causation].
Conversely, an edge marked as $A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
Finally, when an edge is represented as $A \arrowarrow B$, it implies that neither $A$ nor $B$ is an ancestor of the other, suggesting the presence of a latent confounder influencing both variables.

The algorithms also differ in how they detect cycles and represent cyclic relationships in their resulting graphs.
In CCI's MAAG, $A$ \textemdash\ $B$ indicates that $A$ is an ancestor of $B$, and simultaneously $B$ is an ancestor of $A$, referring to a cyclic relationship between $A \stackedarrows B$. 
FCI, on the other hand, identifies potential cycles more subtly. In its PAG, fully-connected nodes with circle endpoints ($\circirc$) may suggest the presence of cyclic structures.
FCI, therefore, provides a sufficient condition to distinguish variables that are not part of a cycle, offering a more nuanced approach to handling cyclic relationships [@mooij2020constraint].

Table 1 highlights further differences in the assumptions underlying these algorithms. CCI operates under the assumption of a linear system, whereas FCI, particularly when used to infer cyclic relationships, assumes a non-linear system without selection bias and adheres to more general faithfulness and Markov conditions (i.e., *$\sigma$-separation* and *$\sigma$-faithfulness* setting) [@forre2018constraint].
When the respective assumptions of each algorithm are met, their inferred cyclic relationships align with those illustrated in @fig-examplegraphs.

While CCI's MAAG provides an explicit representation of cyclic relationships, it has certain theoretical limitations, as the MAAG may not always fully preserve d-separation relations from the original graph, $\mathcal{G}$ [@strobl2019]. Given these considerations, we examine and compare results from both algorithms, placing more weight on the FCI results, which are presented in the main results section, while discussing CCI results where relevant, with detailed findings provided in the [Appendix](#sec-appendix).
A full discussion of causal discovery concepts and algorithmic details is beyond the scope of this paper; however, readers seeking a more in-depth understanding can refer to @park2024discovering for a comprehensive exploration of these methods and their applications.

@@

The graph produced by the PC algorithm is a *CPDAG* (completed partially directed acyclic graph), where directed edges ($A$ → $B$) indicate that $A$ is a direct cause (parent) of $B$. Unlike FCI and CCI, the CPDAG does not include circle symbols. Instead, when the PC algorithm cannot determine the direction of causality, it represents this uncertainty with bidirectional arrows. 
While the PC algorithm serves as a useful reference, its strict assumptions — acyclicity and the absence of latent confounders—limit its applicability in more complex scenarios. For this reason, our primary focus remains on the results obtained from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.

One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear. To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS. Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random].
To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features. This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate]. For a more detailed explanation of RCoT, refer to @sec-rcot.

## Analysis{#sec-analysis}

We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity.
(2) Investigating the relationship between individual depressive symptoms and individual precariousness factors. (3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors).
Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity. However, this aggregated approach may overlook granular symptom-level relationships.

By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors. This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa. However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests.

Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions. This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately. This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis.
By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.

<!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions. -->

Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping. For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds. If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction.

The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms. Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm. 
For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test). This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network. Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred.
Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings.

<!-- [^2] -->
To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups. First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples. Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry.

Since we have an even number of experimental setups, ties may occasionally occur. In such cases, we explicitly denote them in the resulting graph using dashed lines. To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case.

By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately.
This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression. See @fig-workflow for an overview of the analysis workflow.

<!-- [^2]: -->
<!-- We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.01 using RCoT CI test. This approach preserved the complexity of symptom interactions, enabling unrestricted estimation of causal directions both among symptoms (*symptom --- symptom* interactions) and between symptoms and precarity factors. -->

<!-- [^3]: -->
<!-- The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix]. -->

@@

# Results
## Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum). Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role. While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.

The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$. For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.  In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails. The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.

Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder. Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional. Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin. While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression. Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI. Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.

<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.   Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou). Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->

<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*. -->

<!-- Graph (b), derived solely from the nonparametric RCoT method. Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.  -->
<!-- The relationship between stressors and depression remains unresolved between the two algorithms. However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder. -->
<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms. -->

<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways. -->

@@

## Individual depression symptom

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.

The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges. As before, dashed edges denote ties, with specific ties marked by bold horizontal bars. 
In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities. Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type. However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms. Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).

Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors. This suggests that these causal links may be less stable across different conditions. Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.

This discrepancy likely arises from a combination of factors. 
First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery. 
With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies. Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency. RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies. However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings. 
In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful. 
As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead. This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it. This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around. Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable relationships. Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases. Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin). The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.

Among depressive symptoms, dep, glt, and slp appear to be the most connected to precariousness factors, while P.soc has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, slp, sui, and anh emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, con, mot, dep, and glt, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, sleep disturbance (slp) emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network. This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity. On the other hand, social precarity (P.soc) serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.

Finally, the results from CCI are largely consistent with those from FCI, but with some key differences. CCI tends to favor arrowheads more frequently, resulting in a greater number of bidirectional edges, which suggests a higher involvement of latent confounders. Additionally, CCI exhibits more tie situations, though, similar to FCI, most ties occur between absence and arrowhead edges, reflecting the discrepancy between the Gaussian CI test and RCoT—where the Gaussian test favors arrowheads, while RCoT more often suggests the absence of an edge. For a detailed visualization of the CCI-derived graph and proportion matrix, see @fig-sym-cci.

<!-- <!-- Centrality and Connectivity --> 
<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points. Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network. Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors. These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes. -->
<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs. As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.  -->
